Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-08-03 appendix.tex
address appendix comments/lemma 3 question

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
%EM perhaps a section reference for this big book?
Using the notation in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
Next we show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
Similarly, for site split $\{1,3\}$,
\[
    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\]
which also is invariant to exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values, regardless of what we do with $y_1$ and $y_2$.
We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms, again independent of what happens to $y_1$ and $y_2$ (Table~\ref{tab:farris_likelihoods}).
Indeed, the corresponding possibilities for the conditional likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which shows the likelihood remains unchanged if $x_1$ and $x_2$ are swapped.
%EM does this one need a different explanation now?
%das: wrote out in more detail; I realize now I maybe took some of these intermediate steps for granted. Effectively we want to look at exchanging xs and ys together as well as separately, because they result in different arguments of symmetry, but still don't change the likelihood value.
The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ independent of what we do with $x_1$ and $x_2$.

For site splits $\{1,2\}$ and $\{2,3\}$ we see the following.
By exchanging only $x_1$ with $x_2$, the generating probabilities and conditional likelihood values swap between these two site splits.
The same is true of the generating probabilities if we exchange only $y_1$ and $y_2$, except, for the case of the conditional likelihood values, we look at the complement of the most likely ancestral state split.
In other words, the function value for the conditional likelihood also swaps between site splits $\{1,2\}$ and $\{2,3\}$, though the most likely ancestral state split is different.
Now, if we exchange both $x_1$ with $x_2$ and $y_1$ with $y_2$, we see these generating probabilities remain unchanged.
Again, for the conditional likelihood values, we look at the complement of the most likely ancestral state split and see these values also remain unchanged.

@@

We now focus on the conditional log-likelihood.
Assume $\emptyset$ is the most likely ancestral state split for all site splits.
In this case, the conditional log-likelihood has no ambiguity, and has terms given as the log of those in Table~\ref{tab:likelihoods-restricted}.
We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
The conditional log-likelihood
\[
\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t),
\]
using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
\begin{align*}
\ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + C_{x_1}(x_2,y_1,y_2,w).
\end{align*}
where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$.
The derivative with respect to $x_1$ is
\[
\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
\]
which is negative for all $x_1 \ge x^*$.
This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \text{and}\ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

>>>>

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
%EM perhaps a section reference for this big book?
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
Next we show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
Similarly, for site split $\{1,3\}$,
\[
    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\]
which also is invariant to exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values, regardless of what we do with $y_1$ and $y_2$.
We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms, again independent of what happens to $y_1$ and $y_2$ (Table~\ref{tab:farris_likelihoods}).
Indeed, the corresponding possibilities for the conditional likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which shows the likelihood remains unchanged if $x_1$ and $x_2$ are swapped.

%EM does this one need a different explanation now?
%das: wrote out in more detail; I realize now I maybe took some of these intermediate steps for granted. Effectively we want to look at exchanging xs and ys together as well as separately, because they result in different arguments of symmetry, but still don't change the likelihood value.
%das: okay, even gorier detail---I think it's all correct now.
For site splits $\{2\}$ and $\{1,2,3\}$, exchanging $y_1$ and $y_2$ swaps the values of the generating probabilities, independent of what happens to $x_1$ and $x_2$.
In the case of the conditional likelihood values, we see that the values for these site splits swap as well, though, we look at the complement of the most likely ancestral state split.
In other words, the function value for the conditional likelihood also swaps between site splits $\{2\}$ and $\{1,2,3\}$, though the most likely ancestral state split is different.
Indeed,
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{2\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w);\\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{2\}, \tau_1, t)     &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{2\}, \tau_1, t)     &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{2\}, \tau_1, t)   &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w);
\end{align*}
for site split $\{2\}$ and
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2,3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w);\\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)     &= \frac{1}{32}(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)     &= \frac{1}{32}(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w);\\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2,3\}, \tau_1, t)   &= \frac{1}{32}(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w);
\end{align*}
for site split $\{1,2,3\}$, which shows the likelihood remains unchanged if $y_1$ and $y_2$ are swapped.

For site splits $\{1,2\}$ and $\{2,3\}$ we see the following.
By exchanging only $x_1$ with $x_2$, the generating probabilities and conditional likelihood values swap between these two site splits.
The same is true of the generating probabilities if we exchange only $y_1$ and $y_2$, except, for the case of the conditional likelihood values, we again look at the complement of the most likely ancestral state split as in the case of splits $\{2\}$ and $\{1,2,3\}$.
Now, if we exchange both $x_1$ with $x_2$ and $y_1$ with $y_2$, we see these generating probabilities remain unchanged, and, for the conditional likelihood values, we look at the complement of the most likely ancestral state split and see these values also remain unchanged.

@@

We now focus on the conditional log-likelihood.
Assume $\emptyset$ is the most likely ancestral state split for all site splits.
In this case, the conditional log-likelihood has no ambiguity, and has terms given as the log of those in Table~\ref{tab:likelihoods-restricted}.
We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
The conditional log-likelihood
\[
\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t),
\]
using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
\begin{align*}
\ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + C_{x_1}(x_2,y_1,y_2,w).
\end{align*}
where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$.
The derivative with respect to $x_1$ is
\[
\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2},
\]
which is negative for all $x_1 \ge x^*$.
This same process applied to $x_2$ and $y_1$ shows the gradient of the full likelihood is negative for $x_2 \ge x^*$ and $y_1 \ge y^*$.

The variables $y_2$ and $w$ require the following.
Consider the conditional log-likelihood with respect to $w$
\[
\ell^C(w; \boldsymbol\xi_0) := \log(1+w) + C_{w}(x_1,y_1,x_2,y_2)
\]
with derivative with respect to $w$
\[
\frac{d}{dw} \ell^C(w; \boldsymbol\xi_0) = \frac{1}{1+w},
\]
which is positive for all $w \ge 0$.
As the gradient of the marginal log-likelihood is a decreasing function in $w$ and has a zero crossing at $y^*$, the addition of the derivative of the conditional log-likelihood makes the gradient of the full log-likelihood positive for all $w \le y^*$.
The same argument holds for $y_2$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \text{and}\ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}
